Venn diagrams may indicate erroneous statistical reasoning in transcriptomics.

### 2022-04-05 mk5/manuscript.rmd
final update

Here we show that this procedure is statistically incorrect and that the
interaction between treatment and group should be tested instead. Moreover,
we show that gene set enrichment analysis applied to such incorrectly
defined genes group-specific genes may result in misleading artifacts.
Due to the presence of false negatives, genes significant in one, but not
the other group are enriched in gene sets which which correspond to the
overall effect of the treatment.  Thus, the results appear related to the
problem at hand, but do not reflect the group-specific effect of a
treatment. A literature search revealed that more than a quarter of papers
which used a Venn diagram to illustrate the results of separate
differential analysis  have also applied this incorrect reasoning.

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
```

@@

```{r fig1, fig.width=15,fig.height=5.5,dpi=300,dev="png"}
p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
                         G2=res$g2 %>% filter(DEG) %>% pull(ENSEMBL)),
             show_percentage=FALSE) +
      ggtitle("Differentially expressed genes")
p2 <- ggvenn(list(G1=gsea_res$g1 %>% pull(ID),
                         G2=gsea_res$g2 %>% pull(ID)),
              show_percentage=FALSE) +
      ggtitle("Enriched GO terms")

@@

```{r fig2, fig.width=10, fig.height=5.5, dpi=300,dev="png"}
#id    <- gsub(" ", "_", toupper("response to interferon beta"))
#gstitle    <- "GOBP_TOLL_LIKE_RECEPTOR_4_SIGNALING_PATHWAY"
gstitle    <- "GOBP_INTERLEUKIN_7_MEDIATED_SIGNALING_PATHWAY"
gsid  <- gsea_res$g2 %>% filter(grepl(gstitle, Title)) %>% pull(ID)
gsgenes <- mset$MODULES2GENES[[gsid]]
sel   <- res.merged %>% filter(SYMBOL %in% gsgenes) %>% arrange(pvalue.g2) %>% 
  filter(DEG.g2 & !DEG.g1) %>%
  dplyr::slice(1:4) %>% pull(ENSEMBL)

@@

```{r fig3,dpi=300,dev="png"}
tmp <- merge(res.merged, res$full, by="ENSEMBL")
tmp <- tmp[ , c("DEG.g1", "DEG.g2", "DEG") ]
colnames(tmp) <- c("G1", "G2", "Full dataset")
plot(euler(tmp), quantities=list(type=c("counts", "percent")))
```

@@

This manuscript has been written in Rmarkdown [@xie2018r].  
All statistical calculations required to replicate the findings and figures
are contained in the source Rmarkdown file.  The Rmarkdown file, along with
additional files required to recreate this manuscript as well as the
results
of literature survey have been uploaded to
[https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams).

>>>>

Here we show that this procedure is statistically incorrect and that the
interaction between treatment and group should be tested instead. Moreover,
we show that gene set enrichment analysis applied to such incorrectly
defined genes group-specific genes may result in misleading artifacts.
Due to the presence of false negatives, genes significant in one, but not
the other group are enriched in gene sets which correspond to the
overall effect of the treatment.  Thus, the results appear related to the
problem at hand, but do not reflect the group-specific effect of a
treatment. A literature search revealed that more than a quarter of papers
which used a Venn diagram to illustrate the results of separate
differential analysis  have also applied this incorrect reasoning.

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide",dev="jpeg")
```

@@

```{r fig1, fig.width=15,fig.height=5.5,dpi=300,dev="jpeg"}
p1 <- ggvenn(list(G1=res$g1 %>% filter(DEG) %>% pull(ENSEMBL),
                         G2=res$g2 %>% filter(DEG) %>% pull(ENSEMBL)),
             show_percentage=FALSE) +
      labs(title="Differentially expressed genes") +
      ylim(-1, 1.4)
p2 <- ggvenn(list(G1=gsea_res$g1 %>% pull(ID),
                         G2=gsea_res$g2 %>% pull(ID)),
              show_percentage=FALSE) +
      labs(title="Enriched GO terms") +
      ylim(-1, 1.4)

@@

```{r fig2, fig.width=10, fig.height=5.5, dpi=300,dev="jpeg"}
#id    <- gsub(" ", "_", toupper("response to interferon beta"))
#gstitle    <- "GOBP_TOLL_LIKE_RECEPTOR_4_SIGNALING_PATHWAY"
gstitle    <- "GOBP_INTERLEUKIN_7_MEDIATED_SIGNALING_PATHWAY"
gsid  <- gsea_res$g2 %>% filter(grepl(gstitle, Title)) %>% pull(ID)
gsgenes <- mset$MODULES2GENES[[gsid]]
sel   <- res.merged %>% filter(SYMBOL %in% gsgenes) %>% arrange(pvalue.g2) %>% 
  filter(DEG.g2 & !DEG.g1) %>%
  dplyr::slice(1:4) %>% pull(ENSEMBL)

@@

```{r fig3,dpi=300,dev="jpeg"}
tmp <- merge(res.merged, res$full, by="ENSEMBL")
tmp <- tmp[ , c("DEG.g1", "DEG.g2", "DEG") ]
colnames(tmp) <- c("G1", "G2", "Full dataset")
plot(euler(tmp), quantities=list(type=c("counts", "percent")))
```

@@

This manuscript has been written in R markdown [@xie2018r].  
All statistical calculations required to replicate the findings and figures
are contained in the source R markdown file.  The R markdown file, along with
additional files required to recreate this manuscript as well as the
results
of literature survey have been uploaded to
[https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams).
